---
title: "ECO395M Homework3"
author: "Evan Aldrich, Chenxin Zhu, Somin Lee"
date: "2024-04-01"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(ggplot2)
library(modelr)
library(mosaic)
library(foreach)
library(dplyr)
library(rpart)
library(rpart.plot)
library(rsample) 
library(randomForest)
library(lubridate)
library(modelr)
library(gbm)
library(pdp)
library(ggmap)
library(osmdata)
library(tidyr)
library(caret)

```

# Q1: What causes what?

## 1
We can’t get causation by regressing one equation. For example, there can be a third variable that causes crime to decline and police to increase. In that case, we estimate a negative coefficient when we regress crime on police, but it is due to the third variable, not because the police cause crime to decrease. Also crime and police are endogenous, very often the decision for the amount of funding to go to the police department or how many to hire is partially based on the crime level in the city. This third variable would likely be exogenous but act through police to result in more or less crime. 

## 2
High alert day is a instrument variable to police because it is highly correlated with police(=instrument relevance) but not affected others such as the number of people on the street. (They checked there is no ridership decline in high alert day, instrument exogeneity). In the table, model(1) represents that the number of crimes in DC decline by 7.3 in high alert day, and model(2) shows that the number of crimes diminish by 6.0 holding all else fixed(it means that controlling for the ridership). Both of these effects are significant at the 5% level helping us get a better idea of a causal relationship.

## 3
Metro ridership represents a proxy for the potential number of people who can be targeted by crime. If it is not controlled for, we would be omitting a variable and would not be able to tell whether the number of crimes declines due to the police or due to a decrease in the number of people on the street who can be victims of crime.

## 4
In the model, the table shows the result of the regression the daily total number of crimes on 3 explanatory variables. The explanatory variables are dummy variables of High alert(or not), District 1(or not), and one continuous variable log(ridership), and the model use the interaction term between 2 dummy variables and log(ridership). When high alert(=1), the number of crimes decrease by about 2.6 in District 1 compared to other district, holding all else fixed. But the result of it in other districts, -0.57, is not statistically significant. Therefore, the number of crimes in district 1 in High alert day is 2.61 lower than that of other districts, holding all else fixed. This means that a high alert may have a greater effect in stopping crime in district one than the other districts.

# Q2: Tree modeling : Dengue cases

For this question we predict dengue cases (not log cases). We split the data into a training set that accounts for 80% of the observations and a testing set that is the remaining 20% of the observations. We calculate our predictions on the testing set for the CART, random forest, and gradient-boosted trees and compare the out of sample RMSE and find that the gradient boosted tree has the smallest out of sample RMSE in predicting dengue cases.

## 1: CART

We begin with the CART model, we provide below the full tree in which we include all variables in the data set.We have the tree set so the smallest split has at least 10 observations.

```{r pressure, echo=FALSE}

set.seed(125)
dengue = read.csv("dengue.csv")
dengue=na.omit(dengue)
dengue$city = factor(dengue$city)
dengue$season = factor(dengue$season)

dengue_split = initial_split(dengue, prop=0.8)
dengue_train = training(dengue_split)
dengue_test = testing(dengue_split)

tree_cart_big=rpart(total_cases ~ . , data=dengue_train, control=rpart.control(cp=0.0015, minsplit=30)) 
rpart.plot(tree_cart_big, digits=-5, type=4, extra=1)
a= rmse(tree_cart_big, dengue_test)
```

Let's prune our tree at the 1 standard error complexity level. This is the application of the 1SE rule to pick a tree that is simple but whose performance is not discernibly different from the best performing model. Below is this tree.

```{r}
prune_1se = function(my_tree) {
  out = as.data.frame(my_tree$cptable)
  thresh = min(out$xerror + out$xstd)
  cp_opt = max(out$CP[out$xerror <= thresh])
  prune(my_tree, cp=cp_opt)
}

tree_cart_prune1se = prune_1se(tree_cart_big)
rpart.plot(tree_cart_prune1se, digits=-5, type=4, extra=1)
b= rmse(tree_cart_prune1se, dengue_test)

```

Although this tree is much simpler tree than the one previously, with only the first split, the tree_cart_big is the best cart model compared to pruned one. As we are more interested in performance than interpretability, we will use the big tree in our comparisons as the RMSE for the tree_cart_big is less than that of the CART pruned by 1 standard error. It is important to note that the minimum air temperature appears correlate strongly with the number of cases.  

## 2: random forest 
Next, we provide some information on our random forest prediction results. Random forests are effective, fast and require little or no tuning via CV, and we found that the default settings do well for our predictions. Below, we provide the variable importance plot showing when we leave out a variable, how much does that increase mean squared error.

```{r}
tree_rforests = randomForest(total_cases ~ . , data=dengue_train, important=TRUE)
c= rmse(tree_rforests, dengue_test)
varImpPlot(tree_rforests)
```

Here we see that the largest increase in mean squared error would be achieved if we left out minimum air temperature and followed very closely by dew_point_temperature which we also see as a main driving point in the CART tree previously. The variable importance plot is made using the training data but the RMSE was calculated using the testing data is is shown below the next section about the gradient boosted tree.


## 3: gradient-boosted tree

Next we find the predictions of a gradient boosted tree. Below we present the error curve that was used to find the amount of boosting steps. We initially set the depth of of each tree to 4, using 500 trees (like the in-class example), the shrinkage factor at it's typical 0.05. 

```{r}
set.seed(124)
tree_boosted = gbm(total_cases ~ . , data=dengue_train,interaction.depth=4, n.trees=500, shrinkage=.05, cv.folds = 5) 
gbm.perf(tree_boosted) #error curve
d= rmse(tree_boosted,dengue_test)
```

The error curve above is showing as a function of iteration number, how the boosted trees are doing. 


## 4: compare the models
We compile the root mean squared errors of the three methods which are provided below.

```{r}
rmse_compare = c(a,c,d)%>%as.data.frame()
rownames(rmse_compare) <- c("cart_big","random forest","boosted")
colnames(rmse_compare) <- "RMSE"
rmse_compare
```

We see that the out of sample root mean square error of the boosted trees is bit better than the other two. This agrees with our discussions in class that if we construct the ensemble of trees in such a way that the later we get in the tree, the more aware we are in the error of the earlier trees that we can do better. As we are using count outcomes, and making these trees assumes that the count outcomes are Gaussian, which may not always be accurate.


## 5: partial plot

Below we now provide the partial dependence plots for specific humidity, precipitation amount, and minimum air temperature (first split in CART and first in the random forest's variable importance plot) on the gradient boosted machine.

```{r}

pdp_humid_gbm <- partial(tree_boosted, pred.var = "specific_humidity", grid.resolution = 20, n.trees = 500)
plot(pdp_humid_gbm, type = "l", lwd = 2, main = "Partial Dependence of Specific Humidity on Dengue Cases", ylab = "Predicted Cases")

pdp_prec_gbm <- partial(tree_boosted, pred.var = "precipitation_amt", grid.resolution = 20, n.trees = 500)
plot(pdp_prec_gbm, type = "l", lwd = 2, main = "Partial Dependence of Precipitation (in ml) on Dengue Cases", ylab = "Predicted Cases")

pdp_temp_gbm <- partial(tree_boosted, pred.var = "min_air_temp_k", grid.resolution = 20, n.trees = 500)
plot(pdp_temp_gbm, type = "l", lwd = 2, main = "Partial Dependence of Minimum Temp (in K) on Dengue Cases", ylab = "Predicted Cases")

```

The plots above are attempting to show the relationship between the amount of cases of dengue fever, and the aforementioned feature variables in the plot titles. It's attempting to take into account the joint effect of all of the other features in the problem. We see that when the humidity plot that cases are quite low and constant until we get to about 18 in specific humidity then cases increase very rapidly then appear to level off. We see on the precipitation plot that the cases are a bit higher when there's no rain than as it rains a little bit the cases seem to decrease until about 90ml of rain the cases increase again and level off about 200ml of rain per week and beyond. Lastly, minimum air temperature has a very similar shape to the humidity graph as specific humidity and air temperature are likely positively correlated. We see similarly. We see that when temperatures are low, cases are low until we get beyond 296 K or about 73 degrees Fahrenheit, cases increase rapidly and don't appear to level off in the values in the testing data. It appears from these solutions that in warm tropical environments, we can expect more misquotes to be about and infect people with this fever.


# Q3: Predictive model building: green certification

For our solution to this problem, we develop a CART solution, a linear model solution, and a gradient boosted solution. We find that the gradient boosted solution has the least RMSE and we provide the partial dependence plot for the to capture how having a green certified property predicts a increase in the revenue per square foot per calender year for that property.

## Data Preprocessing

The initial step involved cleaning the data set to handle missing values and create new features. Notably, the green_certified variable was engineered to indicate whether a property is green-certified by either LEED or EnergyStar. We also split up the data into a training and testing set wit han 80% split for the training set. We also define the revenue variable as the product of rent and leasing rate. This is what the model will be predicting.

We then create the cart, gbm, and lm models using the training data. We leave out the rent and leasing rate in predicting revenue as those are both the components of the and would result in a colinearity issue. For the GBM, similar to how we compiled the GBM previously in class and this assignment, we set the number of trees to 500, and the interaction depth to 4. We are using a Gaussian distribution for the revenue as its not unlikely that there exists a average property revenue and about equally as many property either earning above or below that average like the Gaussian distribution. 

```{r, message=FALSE, warning = FALSE}

green = read_csv('greenbuildings.csv', show_col_types = FALSE)
green$revenue = green$Rent *green$leasing_rate
green$green_cert <- with(green, ifelse(LEED == 1 | Energystar == 1, 1, 0))

green_split = initial_split(green, prop=0.8)
green_train = training(green_split)
green_test  = testing(green_split)

# LEED and EnergyStar collapse them into a single "green certified" category
#green_train = green_train %>%
#  mutate(green_certified = ifelse(LEED == 1 | Energystar == 1, 1, 0))

#green_test = green_test %>%
#  mutate(green_certified = ifelse(LEED == 1 | Energystar == 1, 1, 0))

```

```{r, message=FALSE, warning = FALSE}
# Model training 
# use all available variables in green_train as predictors except for Rent and leasing_rate
# a.cart model
cart_model <- rpart(revenue ~ . - Rent - leasing_rate, data = green_train, method = "anova")

# b.Gradient Boosting Model
gbm_model <- gbm(revenue ~ . - Rent - leasing_rate, data = green_train, distribution = "gaussian", n.trees = 500, interaction.depth = 4)

# c.Linear Model
lm_model <- lm(revenue ~ . - Rent - leasing_rate, data = green_train)

```

The analysis employed three distinct modeling approaches:CART: Served as a foundation model to establish baseline performance metrics. GBM: An advanced ensemble learning technique selected for its ability to iteratively correct errors and handle complex interactions. Linear Model (LM): A conventional approach providing a benchmark for performance comparison. After training the models, we develop predictions using the testing data and calculate the rmse for the out of sample data and provide the results below. 

```{r, message=FALSE, warning = FALSE}
# Model evaluation 
predictions_cart <- predict(cart_model, newdata = green_test)
predictions_gbm <- predict(gbm_model, newdata = green_test, n.trees = 500)
predictions_lm <- predict(lm_model, newdata = green_test)

# Calculate RMSE for each model
rmse_cart <- sqrt(mean((green_test$revenue - predictions_cart)^2))
rmse_gbm <- sqrt(mean((green_test$revenue - predictions_gbm)^2))
rmse_lm <- sqrt(mean((predictions_lm - green_test$revenue)^2, na.rm = TRUE))

print(paste("CART RMSE:", rmse_cart))
print(paste("Gradient Boosting RMSE:", rmse_gbm))
print(paste("LM RMSE:", rmse_lm))

```

We see again in this problem that the RMSE for the gradient boosted trees is quite a bit lower and therefore the strongest of the three models for predicting rental income per square foot accounting for capacity size (our revenue variable mentioned previously). Using the gradient boosed model we provide the partial dependence plot for green certification on revenue. The plot below is attempting to show the relationship between revenue the property generated, and feature variable that is a 1 if they are green certified or 0 if they are not green certified. It's attempting to take into account the joint effect of all of the other features in the problem as well.

```{r}
pdp_LEED_gbm <- partial(gbm_model, pred.var = "LEED", grid.resolution = 20, n.trees = 500)
plot(pdp_LEED_gbm, type = "l", lwd = 2, main = "Partial Dependence of Green Certification on Revenue", ylab = "Predicted Revenue", xlab = "Green Certification")

```

We see that there is a large increase right about 0.5 that increases predicted revenue by over 200. This split at 0,5 is due to the indicatior variable being either a zero or a one and the large increase suggests that there is likely a positive impact on expected rental income per square foot in absolute terms associated with green certification holding everything else constant. 


# Q4: Predictive model building: California housing

For our answer to this question, we use ggmap and begin by registering an API key with stadia maps to provide a plot of the data points using longitude and latitude on a map surface. After collecting the map image from stadia maps, we plot the original data points on the map of California. We then create a CART, random forest, and a gradient boosted ensemble model to see whic hbest predicts home prices on our testing set. Like the other problems above, we see that the gradient boosed tree provides the lowest out of sample root mean square error. We then plot these best predictions on the same map of california and also plot the errors in absolute terms (predicts - actual) on a map.  

```{r, message=FALSE, warning = FALSE}
register_stadiamaps(key = "c319f1a1-6fbf-4ce5-ab21-37ecfe0d5643", write = TRUE)
bbox <- c(left = -125.6509, bottom = 31.995, right = -112.883, top = 42.7249)
CAmap = get_stadiamap(bbox, zoom = 7)
```

## 1: Original data plot

```{r}

housing_data = read.csv('CAhousing.csv')

ggmap(CAmap) + 
  geom_point(aes(x = longitude, y = latitude, color = medianHouseValue), 
             data = housing_data) +
  labs(x = 'Longitude', y = 'Latitude', title = 'California Housing Values', subtitle = '') 
```

In the original data plot we see that the most expensive houses in California are very much grouped together along the coast as shown by the light blue. this is mostly around the San Francisco area in the central west and the Los Angeles/San Diego regions in the southern parts of California. there is also a bight blue spot by Carson City Nevada. By the color scale we see that most of the housing costs outside of these blue regions are much cheaper and more homogeneous. The darker blue cheaper homes are mostly more inland than the more expensive regions.    

### Data Preprocessing

In the data preprocessing, we split our data set into training and testing with the usual 80% split, and create 2 new variables.We added Bedrooms per household which is an interaction between total bedrooms and 1/households as well as rooms per household which is an interaction between total rooms and 1/households. Given that total rooms and bedrooms are aggregate counts, we normalized these by the number of households in each tract to derive more meaningful features, roomsPerHousehold and bedroomsPerHousehold can be seen as a standardization.   

```{r}

housing_data$bedroomsPerHousehold = housing_data$totalBedrooms /housing_data$households
housing_data$roomsPerHousehold = housing_data$totalRooms / housing_data$households
ca_split = initial_split(housing_data, 0.8)
ca_train = training(ca_split)
ca_test = testing(ca_split)
```


## 2: Model's predictions

We models we created to compare to each other is the CART model using rpart(), a random forest with 500 trees, and a gradienst boosted machine with 500 trees adn a maximum depth of each tree of 4. All three of these models are predicting median house value and are all being trained on the same training data. After the models are trained, we have them each make predictions on the testing data and calculate the out of sample root mean squared errors which are found right below. 

```{r}
# a.cart mode
cart_model = rpart(medianHouseValue ~ ., data = ca_train, method = "anova")
# b.Random Forest Model
rf_model = randomForest(medianHouseValue ~ ., data = ca_train, ntree = 500)
# c.Gradient Boosting Model
gbm_model= gbm(medianHouseValue ~ ., data = ca_train, distribution = "gaussian", n.trees = 500, interaction.depth = 4)

```

# Model Evaluation

```{r}
# Predictions
predictions_cart = predict(cart_model, newdata = ca_test)
predictions_rf = predict(rf_model, newdata = ca_test)
predictions_gbm = predict(gbm_model, newdata = ca_test, n.trees = 500)

# Calculate RMSE
rmse_cart = sqrt(mean((ca_test$medianHouseValue - predictions_cart)^2))
rmse_rf = sqrt(mean((ca_test$medianHouseValue - predictions_rf)^2))
rmse_gbm = sqrt(mean((ca_test$medianHouseValue - predictions_gbm)^2))

# Print RMSE values
print(paste("CART RMSE:", rmse_cart))
print(paste("Random Forest RMSE:", rmse_rf))
print(paste("Gradient Boosting RMSE:", rmse_gbm))

```

Again, like the previous problems in this assignment, the gradient boosted trees have the best out of sample root mean square error so that will be used as out "best" predictive model and plotted to compare the predictions against the original data. We also then plot the error in absolute terms between the predictions of prices and actual prices (predicted - actual) these two are shown in the two plots below. 

## The best performance based on RMS is Gradient Boosting

```{r}
### Answer:The best performance based on RMS is Gradient Boosting
# plot with prediction
# Add Predictions to Dataset
housing_data$predict_medianHouseValue = predictions_gbm
ggmap(CAmap) + 
  geom_point(aes(x = longitude, y = latitude, color = predict_medianHouseValue), 
             data = housing_data) +
  labs(x = 'Longitude', y = 'Latitude', title = "Predicted Median House Value", subtitle = '') 


housing_data$error = housing_data$predict_medianHouseValue - housing_data$medianHouseValue
ggmap(CAmap) + 
  geom_point(aes(x = longitude, y = latitude, color = error), 
             data = housing_data) +
  labs(x = 'Longitude', y = 'Latitude', title = "Gradient Boosting Errors", subtitle = '') 


```

We can see tin both plots that the gradient boosted trees seem to under predict houses that were in the expensive regions in the original data and slightly over predict some of the properties near the areas that are expensive in the original data. The color is slightly misleading though as colors close to the middle imply an accurate prediction but may look like over or under predictions when compared to other colors. Overall the predictions of housing inland seem to be quite accurate and this may serve as a nice tool to predict house prices in California if there a house in a particular region wanting to go on sale.
