---
title: "ECO395M Homework1: Evan Aldrich"
author: "Evan Aldrich"
date: "2024-01-20"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1: Data visualization: flights at ABIA
#### Part A
Make a 2x2 table of conditional probabilities, conditional on the levels of the david.bowie variable. The bottom right entry in your 2x2 table should display P(plays Daft Punk | plays David Bowie).

```{r eval=TRUE, echo=FALSE, message=FALSE, warning = FALSE}
library(tidyverse)
```

```{r eval=TRUE, echo=FALSE}
flights = read.csv('ABIA.csv', header=TRUE)
xtabs(~Cancelled, data=flights)

ggplot(flights) + 
  geom_bar(aes(x = UniqueCarrier)) 



ggplot(flights) + 
  geom_point(aes(x = DepDelay, y = ArrDelay, color=DayOfWeek), width=0.1)

fflights <- flights %>% filter( Dest != 'AUS')
ggplot(fflights) + 
  geom_bar(aes(x = Dest)) + coord_flip() 
```

## Question 2: Wrangling the Olympics
```{r eval=TRUE, echo=FALSE, message=FALSE, warning = FALSE}
top20 = read.csv('olympics_top20.csv', header=TRUE)
```
#### Part A
What is the 95th percentile of heights for female competitors across all Athletics events (i.e., track and field)?
```{r eval=TRUE, echo=FALSE}
female <- top20 %>% filter(sex == "F" & sport == "Athletics")

quantile(female$height, 0.95)



```
#### Part B
Which single women's event had the greatest variability in competitor's heights across the entire history of the Olympics, as measured by the standard deviation?

#### Part C
How has the average age of Olympic swimmers changed over time? Does the trend look different for male swimmers relative to female swimmers? Create a data frame that can allow you to visualize these trends over time, then plot the data with a line graph with separate lines for male and female competitors. Give the plot an informative caption answering the two questions just posed.

## Question 3: K-nearest neighbors: cars

```{r eval=TRUE, echo=FALSE, message=FALSE, warning = FALSE}
sclass = read.csv('sclass.csv', header=TRUE)
```


